Building a continuous benchmarking ecosystem in bioinformatics

### 2024-10-21 manifesto-main.tex
fix
No lineno
Shortening

\linenumbers

@@

In this work, we give our perspective on how to build a high-standard benchmarking ecosystem in the field of bioinformatics, coming primarily from the viewpoint of methodologists (i.e., researchers with computational background working on tools for data). In particular, our case is one of computational scientists embedded in data-rich biology (i.e., computational biology and bioinformatics), where one of the distinguishing factors is rapid technology development (e.g., single-cell assays coupled to high-throughput sequencing, imaging, etc.), which in turn results in a rapid response from the computational community to develop tools, models, and software for the analysis of such emerging datasets. Although our viewpoint is `bioinformatics', there are analogs in adjacent disciplines where methodologists are creating computational tools in a certain scientific context. One critical step in method development is benchmarking (further defined below), whereby either new methods are compared against existing ones, or a set of existing methods are compared in a more neutral way; for these, we will make use of the terms (and abbreviations) `methods-development papers' (MDPs) and `benchmark-only papers' (BOPs), respectively, defined in a recent meta-analysis study \cite{Cao2023-jz}. Notably, there are various ways to structure the running of a benchmarking study: it can be run by a single person or a small group on a laptop (or a server), or it can be organized as a `challenge' or within the duration of a hackathon (groups of people that come together to competitively assess their approaches and/or populate benchmark components).

Our goal here is to lay out general benchmark parameters and definitions, while abstracting towards an ideal \textit{benchmarking system}. This ranges from general benchmark conceptualization to various technical specifications. We will offer some remarks on the philosophy around the benchmarking activity (e.g., gatekeeping), as well as some technical recommendations (e.g., standards).

\section{What is a benchmark?}
%\label{sec:main_section}

To begin our discussion of benchmarking (in bioinformatics), a few definitions are needed. For example, what actually is a benchmark, precisely? Here, a benchmark is thought of as a conceptual entity that roughly encompasses all the components of a study to understand the performance of a set of (computational) methods for a given task (here, the primary context is computational methods, but the concept can have a broader scope, e.g., evaluation of data-generating technologies). A benchmark requires a well-defined task, such as inferring what set of genes are differentially expressed from a certain type of data (e.g., gene expression from RNA sequencing counts) or whether a high-dimensional dataset can be clustered into the `correct' cell types (the definition of this correctness, or ground-truth, should be precisely defined in advance). Given a well-defined task, a benchmark consists of benchmark `components' (such as simulations to generate datasets, preprocessing steps, methods, metrics). Ideally, a benchmarking system is used to organize and orchestrate the running of the benchmark, and create `artifacts' (code snapshots, file outputs to be shared, results tables). Ultimately, the downstream results and interpretations, which may involve rankings and various additional analyses, will be derived from these artifacts. The various elements of code, especially the details on how datasets are preprocessed and how methods are run, should be explicitly available for scrutiny. Such a system should facilitate components to be public, open and allow contributions (e.g. corrections or new components), such that the benchmarks of the future can run more continuously.

At the outset of a benchmark, a benchmark `definition' could be envisioned, which could give a formal specification of the entire set of components (and the pattern of artifacts to be generated). The benchmark definition could even be expressed as a single configuration file that specifies the scope and topology of components to include, details of the repositories (code implementations with versions), the instructions to create software environments (accessible across compute architectures), and the parameters used. A benchmark definition in terms of layers and their corresponding challenges and opportunities within each layer is illustrated in Figure~\ref{fig:layers}.

@@

Benchmarks are useful for a broad set of stakeholders. For a \emph{data analyst}, with an aim to perform a specific type of analysis on a specific dataset, benchmarks are helpful tools to find a suitable method for the task. However, it is important that the benchmarks consider datasets that are similar to the one of interest to the analyst, as method performance often depends on characteristics of the data. Thus, a thorough characterization of the datasets considered in a benchmark is desirable. Given that benchmarks typically evaluate methods using a range of metrics and datasets, not all of which are relevant to all analysts, flexibility in defining the ranking approach and aggregation of metrics is preferable. Such flexibility, which requires access to various artifacts, is typically not part of published benchmarks today \cite{Sonrel2023-te}. However, a well-structured benchmarking system could accommodate flexible filtering and aggregation of metrics and datasets, and additionally provide access to the code and software stack that the analyst would need to apply the selected method to their own dataset.

@@

Benchmark artifacts can be seen as a collection of data files and (perhaps diverse) source code implementations with a pipeline topology (e.g., methods are run on some input data, and method outputs act as input for metrics) that were ideally run as a workflow in some computing environment. Current practices in single-cell RNA-seq benchmarking show that code is frequently open, reusable and versioned; input datasets are also typically available \cite{Sonrel2023-te}. However, the \textit{extensibility} of benchmarking studies is generally very low, as well as the proportion of benchmarks using a formal workflow system. Thus, a formal system to organize components and orchestrate the crossing of datasets, methods, and metrics would largely impact the current landscape.

In an ideal situation, running a benchmark includes executing a workflow, a process for which over 350 workflow languages, platforms or systems  \cite{Wratten2021-el, Amstutz2024-qk} and one standard named Common Workflow Language (CWL) (with multiple implementations) exist \cite{Amstutz2016-vo}. Some workflow languages can export workflow definitions to CWL, hence helping computational FAIR guidelines: that is, bundling and processing data in association to their metadata, tracking new metadata generation, recording provenance, and being accessible \cite{Goble2020-ps}. Both the workflow layout and provenance can be formally defined (Figure~\ref{fig:formalization}).

@@

Benchmarks are often focused on finding the `top' performing method, or on generating ranked lists of methods. While this is valid in a few fields with clearly defined goals, such as image recognition \cite{Khan2018-ot} or protein structure prediction \cite{Jumper2021-oq}, bioinformatics tasks are typically evaluated by multiple metrics, some of which are not well understood \cite{Lutge2021-mt, Reinke2024-dw}. Additionally, ground truths are not always well-defined (e.g., derived with uncertainty from other experimental data), multiple tasks (e.g., represented by different reference datasets) may be evaluated for a single method and several ways to assess performance are possible. Even when accompanied by dataset- and/or metric-specific performance measures, scaling and aggregation of results make it difficult to unpack fine-grained nuances of performance from a final summary table reported as a FunkyHeatmap \cite{funkyheatmap} as in most benchmarks (e.g., \cite{Saelens2019-jy}). Strobl \textit{et al.} push back against the `one method fits all datasets’ philosophy and advocate for highlighting dataset or parameter properties that are predictive of method performance \cite{Jelizarow2010-kr, Boulesteix2010-zj, Strobl2024-lp}. This aligns with what a data analyst wants: to identify the best performers in settings most similar to their problem at hand. This requires not only a thorough benchmark design, e.g., tracking of meaningful dataset metadata \cite{Strobl2024-lp}, but also can be supported by a benchmarking system that facilitates flexible navigation of results (e.g., via an interactive dashboard).

Besides dashboards for more nuanced result exploration \cite{bettr}, multi-criteria decision analysis (MCDA) \cite{Taherdoost2023-wd} has been proposed to guide users through benchmark results and multidimensional scaling (MDS) has been used to differentiate the effect of single datasets or metrics and recognize patterns beyond aggregated performance \cite{Niessl2022-fk}. An extensive analysis and presentation of the results not only helps the users in the long term but also the whole scientific field to identify pain points, the complexity of possible scenarios and to isolate the different use-cases. The first advantage of such a strategy is to force the user to an attentive reading of the performance results so that the correct use-case and relevant methods can be identified. The second, more important, advantage to present benchmark results extensively is to identify the challenges in a field of analysis, i.e., the scenarios, types of datasets, or aspects of performance that are systematically associated with poor results for all current methods. Being public and open to contributions from specialists of different fields, a benchmarking system could gather more reference datasets and methods, and allow the current state of the field to be better grasped.

%As a result, the performance results will tend to better fit to the current state of the field, as opposed to a static benchmark led by a small panel of experts that leads to a smaller, and sometimes (un)intentionally biased view of the field \cite{Boulesteix2010-zj, Weber2019-el}.

@@

In designing a benchmarking system, it may be helpful to specify whether the benchmark is intended to be run `centralized' or `decentralized'. These terms can take different meanings depending on the context, but here the benchmark infrastructure location is intended. In most systems, the code to execute will be tracked in a reference \texttt{git} repository (e.g., on GitHub, GitLab). The workflow execution and the storage of derived datasets, and various stages of results are typically located either in local deployments (e.g., OEB-VRE for \texttt{OpenEBench}), commercial clouds (e.g., AWS for \texttt{OpenProblems}), commercial continuous integration and continuous delivery platforms (e.g., GitHub actions for \texttt{ncbench}) or could be flexible, e.g., on arbitrary infrastructure (i.e., could be a laptop, server or cloud).

Traditionally, benchmarks (whether BOP or MDP) are conducted in a decentralized fashion and typically by a small number of researchers (e.g., for MDP, the new method developer). It is worth mentioning some disadvantages of the decentralized approach. For example, decentralization can be hugely inefficient; researchers studying the same task would collect reference datasets and code snippets to run methods, and implement a bespoke strategy (e.g., shell scripts) to orchestrate the methods run on datasets. Working together would mean sharing not only datasets and code snippets, but also execution environments (software but perhaps also hardware), alignment on parameters to evaluate and eventually sharing, consolidating, and interpreting results. However, because there are no standards of how benchmarks are implemented \cite{Sonrel2023-te}, another large disadvantage of decentralized benchmarks is that, while these components may eventually be openly shared as a product of a research study, the contents of a benchmark are non-interoperable, even across benchmarks of the same task. Furthermore, decentralized benchmarks have been criticized on the basis of `inflating model performance' due to decisions made in preprocessing or parameter tuning \cite{Luecken2024-fk}; such issues could be mitigated by gathering the `wisdom of crowds' that is implicit in running centralized benchmarks.

Arguably, most benchmarks could naturally start out as decentralized, in part as a means to properly define the concept and problem that one wants to work on, but also as a way to generate preliminary results and ideas to solidify the design before `going public' (centralization). Thus, users need a system that can build benchmark components in both a centralized and decentralized fashion, i.e., starting decentralized and promoting a benchmark to centralized when it becomes sufficiently mature. Accordingly, the process of centralization is not cost-free: it implies gatekeeping (so method contributors can authenticate to the system and be given access to computational resources) and adding measures to build trust and transparency.

@@

\bibliographystyle{vancouver}  
\bibliography{clean} % your .bib file

>>>>

%\linenumbers

@@

In this work, we give our perspective on how to build a high-standard benchmarking ecosystem in the field of bioinformatics, coming primarily from the viewpoint of methodologists (i.e., researchers with computational background working on tools for data). Although our viewpoint is `bioinformatics', there are analogs in adjacent disciplines where methodologists are creating computational tools in a certain scientific context. One critical step in method development is benchmarking (further defined below), whereby either new methods are compared against existing ones, or a set of existing methods are compared in a more neutral way; for these, we will make use of the terms (and abbreviations) `methods-development papers' (MDPs) and `benchmark-only papers' (BOPs), respectively, as defined in \cite{Cao2023-jz}. Notably, a benchmarking study can be run by a single person or a small group on a laptop (or a server), or it can be organized as a `challenge' or within the duration of a hackathon.

Our goal here is to lay out an ideal \textit{benchmarking system}, from general benchmark conceptualization to various technical specifications. We offer some remarks on the philosophy around benchmarking (e.g., gatekeeping), as well as some technical recommendations (e.g., standards).

\section*{What is a benchmarking system?}
%\label{sec:main_section}

What actually is a benchmark, precisely? Here, a benchmark is thought of as a conceptual entity that roughly encompasses all the components of a study to understand the performance of a set of (computational) methods for a given task (here, the primary context is computational methods, but the concept can have a broader scope, e.g., evaluation of data-generating technologies). A benchmark requires a well-defined task and a definition of correctness, or ground-truth, precisely defined in advance. Then, a benchmark consists of benchmark `components', such as simulations to generate datasets, preprocessing steps, methods, metrics.

Can benchmarking be systematized? Benchmarks should be orchestrated and run by `benchmarking system's to create `artifacts' (code snapshots, file outputs to be shared, performance outputs).  The various elements of code, especially the details on how datasets are preprocessed and how methods are run, should be explicitly available for scrutiny. Such a system should facilitate components to be public, open and allow contributions (e.g. corrections or new components), such that the benchmarks of the future can run more continuously. At the outset of a benchmark, a benchmark `definition' could be envisioned, which could give a formal specification of the entire set of components (and the pattern of artifacts to be generated). The benchmark definition could even be expressed as a single configuration file that specifies the scope and topology of components to include, details of the repositories (code implementations with versions), the instructions to create software environments (accessible across compute architectures), and the parameters used. A benchmark definition in terms of layers and their corresponding challenges and opportunities within each layer is illustrated in Figure~\ref{fig:layers}.

@@

Benchmarks are useful for a broad set of stakeholders. For a \emph{data analyst}, with an aim to perform a specific type of analysis on a specific dataset, benchmarks are helpful tools to find a suitable method for the task. Hence, it is important that benchmarks include datasets similar to the one of interest to the analyst, as method performance often depends on characteristics of the data. Thus, a thorough characterization of the datasets considered in a benchmark is desirable. Given that benchmarks typically evaluate methods using a range of metrics and datasets, not all of which are relevant to all analysts, flexibility in defining the ranking approach and aggregation of metrics is preferable. Such flexibility, which requires access to various artifacts, is typically not part of published benchmarks today \cite{Sonrel2023-te}. However, a well-structured benchmarking system could accommodate flexible filtering and aggregation of metrics and datasets, and additionally provide access to the code and software stack that the analyst would need to apply the selected method to their own dataset.

@@

Benchmarks can be seen as a collection of data files and (perhaps diverse) source code ideally run as a workflow in some computing environment. Current practices in single-cell RNA-seq benchmarking show that code is frequently open, reusable and versioned; input datasets are also typically available \cite{Sonrel2023-te}. However, the \textit{extensibility} of benchmarking studies is generally very low, as well as the proportion of benchmarks using a formal workflow system. Thus, a formal system to orchestrate benchmarking workflows would largely impact the current landscape.

Over 350 workflow languages, platforms or systems  \cite{Wratten2021-el, Amstutz2024-qk} and one standard named Common Workflow Language (CWL) (with multiple implementations) exist \cite{Amstutz2016-vo}. Some workflow languages can export workflow definitions to CWL, hence helping computational FAIR guidelines: that is, bundling and processing data in association to their metadata, tracking new metadata generation, recording provenance, and being accessible \cite{Goble2020-ps}. For benchmarking, we believe both the workflow layout and provenance can be formally defined (Figure~\ref{fig:formalization}).

@@

Benchmarks are often focused on finding the `top' performing method, or on generating ranked lists of methods. While this is valid in a few fields with clearly defined goals, such as image recognition \cite{Khan2018-ot} or protein structure prediction \cite{Jumper2021-oq}, bioinformatics tasks are typically evaluated by multiple metrics, some of which are not well understood \cite{Lutge2021-mt, Reinke2024-dw}. Additionally, ground truths are not always well-defined (e.g., derived with uncertainty from other experimental data) and multiple tasks (e.g., represented by different reference datasets) may be evaluated for a single method and several ways to assess performance are possible. Even when accompanied by dataset- and/or metric-specific performance measures, scaling and aggregation of results make it difficult to unpack fine-grained nuances of performance from a final summary table reported as a FunkyHeatmap \cite{funkyheatmap} as in most benchmarks (e.g., \cite{Saelens2019-jy}). Strobl \textit{et al.} push back against the `one method fits all datasets’ philosophy and advocate for highlighting dataset or parameter properties that are predictive of method performance \cite{Jelizarow2010-kr, Boulesteix2010-zj, Strobl2024-lp}. This aligns with what a data analyst wants: to identify the best performers in settings most similar to their problem at hand. This requires not only a thorough benchmark design, e.g., tracking of meaningful dataset metadata \cite{Strobl2024-lp}, but also can be supported by a benchmarking system that facilitates flexible navigation of results.

Besides interactive dashboards for more nuanced result exploration \cite{bettr}, multi-criteria decision analysis (MCDA) \cite{Taherdoost2023-wd} has been proposed to guide users through benchmark results and multidimensional scaling (MDS) has been used to differentiate the effect of single datasets or metrics and recognize patterns beyond aggregated performance \cite{Niessl2022-fk}. An extensive analysis and presentation of the results not only helps the users in the long term but also the whole scientific field to identify pain points, the complexity of possible scenarios and to isolate the different use-cases. The first advantage of such a strategy is to force the user to an attentive reading of the performance results so that the correct use-case and relevant methods can be identified. The second, more important, advantage to present benchmark results extensively is to identify the challenges in a field of analysis, i.e., the scenarios, types of datasets, or aspects of performance that are systematically associated with poor results for all current methods. Being public and open to contributions from specialists of different fields, a benchmarking system could gather more reference datasets and methods, and allow the current state of the field to be better grasped.

@@

In designing a benchmarking system, it may be helpful to specify whether the benchmark is intended to be run `centralized' or `decentralized'. These terms can take different meanings depending on the context, but here the benchmark infrastructure location is considered. In most systems, the code to execute will be tracked in a reference \texttt{git} repository (e.g., on GitHub, GitLab). The workflow execution and the storage of derived datasets, and various stages of results are typically located either in local deployments (e.g., OEB-VRE for \texttt{OpenEBench}), commercial clouds (e.g., AWS for \texttt{OpenProblems}), commercial continuous integration and continuous delivery platforms (e.g., GitHub actions for \texttt{ncbench}) or could be flexible, e.g., on arbitrary infrastructure (i.e., could be a laptop, server or cloud; e.g. omnibenchmark \cite{omnibenchmark}).

Traditionally, benchmarks (whether BOP or MDP) are conducted in a decentralized fashion and typically by a small number of researchers (e.g., for MDP, the new method developer). This brings some disadvantages. For example, decentralization can be hugely inefficient; researchers studying the same task would collect reference datasets and code snippets to run methods, and implement a bespoke strategy (e.g., shell scripts) to orchestrate the methods run on datasets. Working together would mean sharing not only datasets and code snippets, but also execution environments (software but perhaps also hardware), alignment on parameters to evaluate and eventually sharing, consolidating, and interpreting results. However, because there are no standards of how benchmarks are implemented \cite{Sonrel2023-te}, another large disadvantage of decentralized benchmarks is that, while these components may eventually be openly shared as a product of a research study, the contents of a benchmark are non-interoperable, even across benchmarks of the same task. Furthermore, decentralized benchmarks have been criticized on the basis of `inflating model performance' due to decisions made in preprocessing or parameter tuning \cite{Luecken2024-fk}; such issues could be mitigated by gathering the `wisdom of crowds' that is implicit in running centralized benchmarks.

Arguably, most benchmarks could start out as decentralized, in part as a means to properly define the concept and problem, but also as a way to generate preliminary results and ideas to solidify the design before `going public' (centralization). Thus, users need a system that can build benchmark components in both a centralized and decentralized fashion, i.e., starting decentralized and promoting a benchmark to centralized when it becomes sufficiently mature. Accordingly, the process of centralization is not cost-free: it implies gatekeeping (so method contributors can authenticate to the system and be given access to computational resources) and adding measures to build trust \textit{via} transparency.

@@

\bibliographystyle{vancouver} % elsarticle-num} % vancouver}  
\bibliography{clean} % your .bib file
